sensi_an_rotations1. Start of sensitivity analysis for rotations.

### 2018-05-02 inst/documents/paper3/paper3_rotations.Rmd
paper3 first draft

  # load simulation outputs for some plots
  # beware that each object is called df_res_all
  # some plots are created by directly running the model
  load("df_res_all_rot_nocost_nomig1000.rda") #nocost nomig  
  df_nocost_nomig <- df_res_all
  load("df_res_all_rot10000.rda") #cost
  df_cost <- df_res_all
  load("df_res_all_rot_mig1000.rda") #mig
  df_mig <- df_res_all

@@

@@

A population genetic model was developed simulating changes over time in the frequencies of resistant and susceptible alleles in response to fitness differences. The model uses a similar approach to that described in [@Levick2017,@South2018] to compare insecticide mixtures and sequences. The implementation is simpler here because for rotations and sequences there is only a need to follow one insecticide at a time. The simpler implementation allows the model to be run for an unlimited number of insecticides. The algebra behind the model is described in the supplementary information.

Sequences : one insecticide is used until resistance threshold frequency (0.5 in this case) is reached, then switch to new insecticide. 
Rotations : use insecticide for set time, switch to other.

In both cases continue until no more insecticides below the resistance threshold remain.

@@

The model was run under four main scenarios :
1. no resistance fitness costs or dispersal link to untreated areas
2. fitness costs of resistance added
3. dispersal link to untreated areas (no costs)
4. costs and dispersal

@@

@@

When there are no resistance fitness costs or dispersal links to untreated refugia resistance frequencies do not decline when an insecticide is not in use. Thus for sequences once resistance thresholds are reached the resistance frequencies remain at that level and the insecticide cannot be re-used. For rotations resistance frequencies step upwards when insecticides are in use and remain at a plateau when they are not.

When there were no costs or dispersal there was no difference between rotations and sequences in the number of generations below the resistance threshold.

@@

In contrast to the previous section 7% of model runs generated an advantage of greater than 20% for rotations over sequences.

@@

@@

1. rotations and sequences the same when no costs or dispersal
2. with resistance costs, 67% of runs still the same for rotations and sequences, but 7% of runs have > 20% advantage for rotations
3. with dispersal, ...

In very rare circumstances rotations can lead to the resistance frequencies for a group of insecticides being kept below resistance thresholds for a very long period of time (and substantially longer than for sequential use). Is it possible that these rare circumstances where parameter values are 'just-right' could be implemented operationally ?

@@

In their effect on the evolution of insecticide resistance, rotations and sequences are most often the same. Infrequently, in less than 10% of model runs set between plausible limits, there are substantial benefits to the rotation strategy over a sequence. There was naver a substantial benefit of a sequence over a rotation. Whether or not insecticide rotations are predicted to offer sizeable benefits over sequences is dependent on coarse and fine scale issues about the nature of insecticide resistance that are yet to be agreed upon. The model shows that resistance costs and/or dispersal from untreated refugia are necessary to provide an evolutionary advantage to rotations. However even when these requirements are satisfied particular combinations of inputs are required to generate the advantage of rotations.

@@

```{r Fig4, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7, fig.width=9, fig.show='hold', fig.align='center', fig.cap="Figure 4. Effect of cost and dispersal on the difference between rotations and sequences"}

  gg_nocost_nomig <- ggplot( df_nocost_nomig, aes(x=rot_minus_seq)) +

             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('no cost or dispersal : rotation generations minus sequence') +
             ylab('count (sqrt scale to show low values)') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  gg_cost <- ggplot( df_cost, aes(x=rot_minus_seq)) +

             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('cost : rotation generations minus sequence') +
             ylab('count (sqrt scale to show low values)') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  gg_mig <- ggplot( df_mig, aes(x=rot_minus_seq)) +

             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('dispersal : rotation generations minus sequence') +
             ylab('count (sqrt scale to show low values)') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  grid.arrange(gg_nocost_nomig, gg_cost, gg_mig, ncol=1)

>>>>

  # load simulation outputs for some plots
  # beware that each object is called df_res_all
  # some plots are created by directly running the model
  load("df_res_all_rot_nocost_nomig1000.rda") #nocost nomig  
  df_nocost_nomig <- df_res_all
  load("df_res_all_rot10000.rda") #cost
  df_cost <- df_res_all
  load("df_res_all_rot_mig10000.rda") #mig 10k
  df_mig <- df_res_all
  load("df_res_all_rot_cost_mig10000.rda") #costmig 10k
  df_cost_mig <- df_res_all

@@

In a rotation the insecticide used is changed on a set time interval irrespective of resistance levels. Sequential use, in contrast, denotes the practice of using an insecticide until resistance to it reaches a certain threshold and then switching to another. Different rotation time intervals can be considered, e.g. in vector control for Malaria, Indoor Residual Spraying may be repeated every year (around 10 anopheles mosquito generations) and Insecticide Treated Nets may be replaced every two years (around 20 generations). Multiples of these intervals could be used as the rotation interval. Multiple generational rotations like this contrast with the situation in agriculture where a rotation is sometimes more tightly defined as having an interval of a single generation [@Sudo2017].

@@

A population genetic model was developed simulating changes over time in the frequencies of resistant and susceptible alleles in response to fitness differences. The model uses a similar approach that has been used to compare insecticide mixtures and sequences [@Levick2017,@South2018]. The implementation is simpler here because for rotations and sequences there is only a need to follow one insecticide at a time. The simpler implementation allows the model to be run for an unlimited number of insecticides. The algebra behind the model is described in the supplementary information.

1. Sequences : one insecticide is used until resistance threshold frequency (0.5 in this case) is reached, then switch to a new insecticide.

2. Rotations : use insecticide for set time, switch to another.

In both cases the choice of the new insecticide to use was made by going through the list of available insecticides and choosing the next one for which resistance was below the threshold. Thus there is the potential to switch back to insecticides that had been used earlier in a model run if resistance had not reached the threshold or if resistance had declined below the threshold during a period when not in use.

In both cases a simulation run continued until no more insecticides remained below their resistance thresholds.

@@

The model was run under four main scenarios :

1. no resistance fitness costs or dispersal link to untreated areas
2. fitness costs of resistance added
3. dispersal link to untreated areas (no costs)
4. costs and dispersal

@@

Example model runs are shown with no costs or dispersal (Fig 1), with just resistanace costs (Fig 2) and with just dispersal from untreated refugis (Fig 3).

@@

When there are no resistance fitness costs or dispersal links to untreated refugia resistance frequencies do not decline when an insecticide is not in use (Fig 1). Thus for sequences once resistance thresholds are reached the resistance frequencies remain at that level and the insecticide cannot be re-used. For rotations resistance frequencies step upwards when insecticides are in use and remain at a plateau when they are not.

When there were no costs or dispersal there was no difference between rotations and sequences in the number of generations below the resistance threshold (Fig 4.1).

@@

In contrast to the previous section 7% of model runs generated an advantage of greater than 20% for rotations over sequences (Fig 4.2).

@@

Dispersal from untreated refugia gives smaller benefits of rotations over sequences and in fewer runs (Fig 4.3).

#### Results4. costs and dispersal

With both costs and dispersal the benefits of rotations are similar to when runs have costs alone (Fig 4.4).

#### Runs where rotations have substantial advantage over sequences
As shown in Fig 4.2 some runs with costs of resistance can show substantial advantages for rotations over sequences. Examples of one such run is shown in Fig 5.

A Partial Rank Correlation analysis identified that of the inputs, dominance of cost and dominance of selection had the greatest influence on the difference between rotations and sequences when resistance costs were included (Fig 6). A combination of high values of dominance of cost and low values for dominance of selection are required to give the greatest advantages to rotations, but do not guarantee this advantage (Fig 7).

@@

1. rotations and sequences the same when no costs or dispersal
2. with resistance costs, 67% of runs the same for rotations and sequences, but 6-7% of runs have > 20% advantage for rotations
3. with dispersal from untreated refugia but no costs, small advantage of rotations but very few runs < 0.1% have >20% advantage.

In rare circumstances, when there are costs of resistance, rotations can lead to the resistance frequencies for a group of insecticides being kept below resistance thresholds for the length of the simulations where sequences run out of insecticides within around 100 generations. Is it possible that these rare circumstances where parameter values are 'just-right' could be implemented operationally ? The problem is that these rare circumstances rely on high costs of resistance, high dominance of these costs and low dominance of the selection for the resistance itself. These parameters are all not easy to measure and there isn't agreement on their likely values [@Kliot2012, @Bourguet2000].

The importance of resistance costs for substantial evolutionary benefits of rotations over sequences points to the importance of the debate over the fitness costs of insecticide resistance e.g. [@Kliot2012].
*extend this discussion of fitness costs*

@@

In their effect on the evolution of insecticide resistance, rotations and sequences are most often the same. If costs of resistance are included, infrequently (less than 7% of model runs set between plausible limits) there are substantial benefits to the rotation strategy over a sequence. There was never a substantial benefit of a sequence over a rotation. Whether or not insecticide rotations are predicted to offer sizeable benefits over sequences is dependent on coarse and fine scale issues about the nature of insecticide resistance that are yet to be agreed upon. The model shows that resistance costs are necessary to provide a substantial evolutionary advantage to rotations. However even when these requirements are satisfied particular combinations of inputs are required to generate the advantage of rotations.

@@

```{r Fig4, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7, fig.width=9, fig.show='hold', fig.align='center', fig.cap="Figure 4. Effect of cost and dispersal on the distribution of differences between rotations and sequences from model runs. The y axis is square root transformed to emphasise the infrequent runs where there is a significant benefit of rotations over sequences. 4.1 when there are no resistance costs or dispersal from untreated refugia, rotations and sequences produce the same results. 4.2, 4.4 when costs are added there are infrequent model runs where rotations last longer below resistance thresholds than sequences. For between 6 and 7% of runs including costs, rotations last more than 20% longer than sequences. 4.3 when dispersal from untreated refugia is included without costs there are even less frequent runs where rotations last longer than sequences (in just 4 out of 10000 runs rotations lasted greater than 20% longer than sequences)."}

  # remove runs where thresholds not reached for seq & rot
  # with cost_mig ~ 65% of runs don't reach thresholds
  # could get this func to output perecnt not reaching
  rem_thresh_not_reached <- function(x) {x %>% filter(gens_seq != max_gen | gens_rot != max_gen)}

  df_nocost_nomig <- rem_thresh_not_reached(df_nocost_nomig)
  df_cost <- rem_thresh_not_reached(df_cost) #6.6
  df_mig <- rem_thresh_not_reached(df_mig) #0.05
  df_cost_mig <- rem_thresh_not_reached(df_cost_mig) #6.9

  # add a column for percent difference between rot & seq
  df_nocost_nomig <- mutate(df_nocost_nomig, pcent_rot_adv = rot_minus_seq * 100/gens_seq)
  df_cost <- mutate(df_cost, pcent_rot_adv = rot_minus_seq * 100/gens_seq)
  df_mig <- mutate(df_mig, pcent_rot_adv = rot_minus_seq * 100/gens_seq)
  df_cost_mig <- mutate(df_cost_mig, pcent_rot_adv = rot_minus_seq * 100/gens_seq)

  # calc % of runs where rotations > 20% better
  pcent_runs_rot_20pc_better <- function(x){ x %>%
                  filter(rot_minus_seq/gens_seq >= 0.2 ) %>%
                  summarise(n=n()) %>% 
                  #unlist() %>%
                  mutate(pcent=n*100/nrow(x)) }

  # pcent_runs_rot_20pc_better(df_nocost_nomig)
  # pcent_runs_rot_20pc_better(df_cost)
  # pcent_runs_rot_20pc_better(df_mig)
  # pcent_runs_rot_20pc_better(df_cost_mig)

  gg_nocost_nomig <- ggplot( df_nocost_nomig, aes(x=pcent_rot_adv)) +

             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('') +
             ylab('count') +
             ggtitle('1. no resistance costs or dispersal') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  gg_cost <- ggplot( df_cost, aes(x=pcent_rot_adv)) +

             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('') +
             ylab('count') +
             ggtitle('2. with costs') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  gg_mig <- ggplot( df_mig, aes(x=pcent_rot_adv)) +

             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('') +
             ylab('count') +
             ggtitle('3. with dispersal refugia') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  #gg_cost_mig <- ggplot( df_cost_mig, aes(x=rot_minus_seq)) + 
  gg_cost_mig <- ggplot( df_cost_mig, aes(x=pcent_rot_adv)) +              
             geom_histogram(binwidth=10) +
             xlim(-20,400) +
             xlab('percent benefit of rotation versus sequence') +
             ylab('count') +
             ggtitle('4. with costs and dispersal') +
             scale_y_sqrt(breaks=c(0,10,20,50,200,500,2000,5000)) + #can't use log scale because anchored at 0
             theme_minimal() +
             theme(panel.grid.minor.y=element_blank())

  grid.arrange(gg_nocost_nomig, gg_cost, gg_mig, gg_cost_mig, ncol=1)

```{r Fig5, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7, fig.width=9, fig.show='hold', fig.align='center', fig.cap="Figure 5. Example of a model run where there is a substantial benefit of rotations over a sequence. For the sequence resistance thresholds are reached for all insecticides in a little over 100 generations or 10 years. For the rotation resistance levels for all insecticides are maintained well below resistance thresholds for the length of the simulation (500 generations or 50 years)."}
# ? are these run ids before or after the runs that don't reach thresholds have been removed ?
# seems OK that they are after other runs removed
#scenarios_to_plot <- c(624,2167,602,4028,5771)
scenarios_to_plot <- c(2167)

for(scenario in scenarios_to_plot)
{
  #cat(scenario,"\n")

  linputs <- as.list( df_cost[scenario, 1:15] )

  # run one scenario
  dfres <- do.call(run_rot, linputs)

  ggrot <- rot_plot_resistance(dfres, plot_refuge=FALSE, logy=TRUE, plot=FALSE)

  #now run as sequence
  linputs$rot_interval <- 0
  dfres <- do.call(run_rot, linputs)

  ggseq <- rot_plot_resistance(dfres, plot_refuge=FALSE, logy=TRUE, plot=FALSE)

  grid.arrange(ggrot, ggseq, ncol=1)

}

```{r Fig6, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7, fig.width=9, fig.show='hold', fig.align='center', fig.cap="Figure 6. Which model inputs have the greatest influence on the difference between rotations and sequences when costs alone are included ? Measured using Partial Rank Correlation Coefficients (PRCC). Dominance of cost and dominance of selection have the greatest influence."}

library(sensitivity)

# pcc(X, y, rank = FALSE, nboot = 0, conf = 0.95)
# 
# Arguments
# X a data frame containing model input variables
# y a vector containing the responses 
# rank logical. If TRUE, the analysis is done on the ranks.
# nboot the number of bootstrap replicates.
# conf the confidence level of the bootstrap confidence intervals.

#same as above
in_cols <- c('cost','dom_cos','dom_sel','rr','expo_hi','male_expo_prop','eff','start_freqs')

    x <- df_cost[,in_cols]
    #y <- df_cost['rot_minus_seq']
    y <- df_cost['pcent_rot_adv']

    pcc_res <- pcc(x, y, rank=TRUE)

    #if you add bootstrap, error bars are added to the default plot & extra columns to the PRCC object
    #pcc_res <- pcc(x, y, rank=TRUE, nboot=100)
    #plot(pcc_res)

    #results are here I can probably rbind them together into a df that I can ggplot
    #pcc_res$PRCC

    to_plot <- pcc_res$PRCC
    #rename column 1 from 'original to PRCC
    names(to_plot)[1] <- 'PRCC'
    to_plot$inputs <- rownames(to_plot)

    axis_text_x <- element_text(angle = 45,hjust = 1, vjust = 1)
    height <- 0.9

    gg <- ggplot( to_plot, aes_string(x='inputs',y='PRCC') ) +     
             geom_point(shape=1, colour='red') +
             theme(axis.text.x = axis_text_x) +
             geom_hline(yintercept = 0, linetype=3) +
             #theme(panel.grid.major.y = element_line(color = "grey", size = 1)) +
             ylim(-1,1) +
             #ggtitle(paste(strat_name[strategy_num])) +
             xlab(NULL)

    plot(gg)

  #to put back in some ref gridlines that are removed by cowplot
  #theme_set(theme_gray())
  #theme_set(theme_bw())

  #plot_grid( plotlist[[1]],plotlist[[2]],plotlist[[3]],plotlist[[4]],ncol=1, rel_heights=c(1,1,1,1.8))  
  #plot_grid( plotlist[[1]],plotlist[[2]],plotlist[[3]],ncol=1, rel_heights=c(1,1,1.6), labels='AUTO')

```{r Fig7, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE, fig.height=7, fig.width=9, fig.show='hold', fig.align='center', fig.cap="Figure 7. Model runs with substantial benefits of rotations over sequences are associated with high dominance of cost and low dominance of selection. Dominace inputs are shown on the x and y axes. Small grey points indicate all model runs. Model runs in which rotations last greater than 20% longer than sequences are shown in colour with darker blue indicating those that have the greatest difference."}

#'cost','dom_cos','dom_sel','rr','expo_hi','male_expo_prop','eff','start_freqs'
#'cost','dom_cos','dom_sel'

y <- 'pcent_rot_adv'

gg <- ggplot(df_cost, aes(x=dom_sel, y=dom_cos)) + 
     geom_point(shape='.', size=1, alpha=0.2, show.legend=FALSE) +
     theme_minimal() +
     #geom_point(data=filter(df_cost, pcent_rot_adv >= 20 ), shape=1, size=1, col='red', alpha=0.8)
     #shape=19 filled circle, 1=open circle
     geom_point(data=filter(df_cost, pcent_rot_adv >= 20), aes(col=pcent_rot_adv), shape=19, size=1, alpha=0.9) +
     viridis::scale_color_viridis(direction=-1) #discrete = TRUE)

plot(gg)

### 2018-05-02 inst/documents/sensi_an_rotations1.Rmd
paper3 first draft

# set random seed so that results can be reproduced
#set.seed(1)
set.seed(2)
max_gen         <- 500
plot_scenarios  <- FALSE
n_scenarios     <- 1000 # try 1000 runs with no cost
#already_ran     <- FALSE
already_ran     <- TRUE

@@

    ## variable inputs ##
    #linputs$cost <-           runif(1, min=0,   max=0.1) 
    # try 1000 runs with no cost
    linputs$cost <-           0 
    linputs$expo_hi <-        runif(1, min=0.4, max=0.9)  #min=0.1, max=0.9
    linputs$male_expo_prop <- runif(1, min=0,   max=1)
    linputs$eff <-            runif(1, min=0.5, max=1) # min=0.3, max=1
    linputs$rr  <-            runif(1, min=0.1, max=0.9)
    linputs$dom_sel <-        runif(1, min=0,   max=1)
    linputs$dom_cos <-        runif(1, min=0,   max=1) 
    linputs$start_freqs <-    runif(1, min=0.005,   max=0.1) #note not log yet   
    linputs$n_insecticides <- sample( 2:5, 1 )  #beware set replace=T if more than 1
    # dispersal (initially with cost set to 0)
    linputs$coverage <-       runif(1, min=0.1, max=0.9)
    linputs$migration <-      runif(1, min=0.1, max=0.9)

>>>>

# set random seed so that results can be reproduced
#set.seed(1)
set.seed(2)
max_gen         <- 500
plot_scenarios  <- FALSE
n_scenarios     <- 10000 # try 1000 runs with no cost
already_ran     <- FALSE
#already_ran     <- TRUE

@@

    ## variable inputs ##
    linputs$cost <-           runif(1, min=0,   max=0.1) 
    #linputs$cost <-           0 
    linputs$expo_hi <-        runif(1, min=0.4, max=0.9)  #min=0.1, max=0.9
    linputs$male_expo_prop <- runif(1, min=0,   max=1)
    linputs$eff <-            runif(1, min=0.5, max=1) # min=0.3, max=1
    linputs$rr  <-            runif(1, min=0.1, max=0.9)
    linputs$dom_sel <-        runif(1, min=0,   max=1)
    linputs$dom_cos <-        runif(1, min=0,   max=1) 
    linputs$start_freqs <-    runif(1, min=0.005,   max=0.1) #note not log yet   
    linputs$n_insecticides <- sample( 2:5, 1 )  #beware set replace=T if more than 1
    # dispersal (initially with cost set to 0)
    linputs$coverage <-       runif(1, min=0.1, max=0.9)
    linputs$migration <-      runif(1, min=0.1, max=0.9)
